---
title: "Practical Machine Learning"
date: "August 23, 2015"
output: html_document
---

This document describes the study done based on the rules imposed by the Practical Machine Learning course provided by Coursera (predmachlearn-031).
The full objective is to identify a predictive machine learning algorithm to classify measurements gotten from movements done with wearables. Full description from the project goals:

> Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

# Data preparation

The first step is to get the data and prepare it for the analysis understading and cleaning it.

```{r, warning=FALSE}
get.pml.files <- function() {
  
  if (! file.exists("pml-training.csv")) {
    download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", "pml-training.csv", method="curl")
  }
  if (!file.exists("pml-testing.csv")) {
    download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", "pml-testing.csv", method="curl")
  }
  
  if (! exists('training')) {
    training <- read.csv('pml-training.csv', 
                         header=TRUE, 
                         na.strings=c("","#DIV/0!","NA")
                         #stringsAsFactors=FALSE
                         )
  }    
  training
}

remove_na_columns <- function(training) {
  require('dplyr')
  
  # Called my attention that many (67) columns has exactly 19216 NAs
  nas <- apply(training, 2, function(x) {sum(is.na(x))})
  colnas <- data.frame(nas)
  colnas['colname'] <- row.names(colnas)
  colnames_nas <- colnas %>% filter(nas > 0) %>% select(colname)
  
  # Get the training without NAs cols
  if (dim(colnames_nas)[1] == 0) {
    training
  }
  else {
    training[, -which(names(training) %in% colnames_nas[['colname']])]
  }
  
  # There are 406 observations with no NAs. 
  # We should research if the obs without NAs can create
  # a better model than doing it with all obs
}
```

406 observations have been identified as complete measurements without NAs in most of the columns. Since the number observations is really high (19622 removing the columns with most of NAs), it could be a good start to train the model. In case that the model accuracy is not good, we can step back and try the 406 observations where almost all columns contain data.

Once the data has been clean, the next is to split it into training and testing to test accuracy later.
```{r, warning=FALSE, error=FALSE, message=FALSE}
set.seed(48293)
training <- remove_na_columns(get.pml.files())

require('caret')

# Create a set of training and testing set to measure error
if (!exists('pmlTraining')) {
  inTrain <- createDataPartition(y=training$classe,p=0.75,list=FALSE)

  pmlTest <- training[-inTrain,]    
  pmlTraining <- training[inTrain,]
  
  # Also identified that there are columns which won't help 
    pmlTraining <- pmlTraining[,-which( names(pmlTraining) %in% 
                                        c('user_name', 'cvtd_timestamp', 'new_window', 'X'))]
}
```

I struggled for a long time in the model accuracy using the training because of include the row number as part of the training
It was key to remove that column (named X).

# Training the models

The first step in training is to identify how many variables (columns) can cover the 95% of the variance in the data. Using PCA we get:
```{r, warning=FALSE, echo=FALSE, message=FALSE}
# Since I wanted to try the caret train function with boosting, I decided to do it in parallel
library(doMC)
registerDoMC(cores = 3)

# although the train method does it for you, let's see how many variables could cover the 90% of variance in the data
preProc <- preProcess(
  pmlTraining[,-which( names(pmlTraining) %in% 
                         c('classe','user_name', 'cvtd_timestamp', 'new_window'))], 
                       method='pca', thresh=0.95)

preProc
```

In order to compare accuracy from different models, below are three models for classifying the data and providing the output.

The three models were done with the PCA default preprocessing based on the information seen above.

Also, bootstrap (default cross-validation method in caret train method) was used adjusting the number of folds to a low number due to the large amount of data in the data set. 

```{r, warning=FALSE, message=FALSE}
# Building tree model
if (!exists('treeModel')) {
  treeModel <- train( classe ~ ., method='rpart', preProcess='pca', data=pmlTraining)
}

if (!exists('rfModel')) {
  # Since the data is large enough to train well our model, the number of folds can be reduced to a
  # small number, for instance, 5 (actually, I've tried with no folds and works very good as well)
  # The number of trees can also be lowered to 50 due to the amount of variables predicting
  rfModel <- train( classe ~ ., method='rf', preProcess='pca', data=pmlTraining, prox=TRUE, ntree=50,
                    trControl = trainControl(number=5, repeats=1))
}

if (!exists('nbModel')) {
  nbModel <- train( classe ~ ., method='nb', preProcess='pca', data=pmlTraining,
                    trControl = trainControl(number=5, repeats=1))
}
```

Once models were built, let's see which is the accuracy for each one

```{r,warning=FALSE, message=FALSE, echo=FALSE}
# Confusion matrix
if (exists('treeModel')) {
  resTree <- predict(treeModel, pmlTest)
  confusionMatrix(resTree, pmlTest$classe)
}

# Confusion matrix
if (exists('rfModel')) {
  resRF <- predict(rfModel, pmlTest)
  confusionMatrix(resRF, pmlTest$classe)
}

# Confusion matrix against NB
if (exists('nbModel')) {
  resNB <- predict(nbModel, pmlTest)
  confusionMatrix(resNB, pmlTest$classe)
}

```

From the above results, the selected model is the random forest.
The accuracy for the tree model was 41% while for the random forst was 97% and Naibe Bayes was 66%
There is no need to test the 406 observations without NA though it could have been a good exercise.

Error accuracy can be seeing in the Sensitivity and Specificity that are really high for the random forest validating the selection.